Polynomial phase estimation by phase unwrapping

### 2012-10-15 paper.tex
edited paper.  Simplified plots.

\begin{aug}
\author{\fnms{Robby G.} \snm{McKilliam}\thanksref{m1}\ead[label=e1]{robby.mckilliam@unisa.edu.au}},
\author{\fnms{Barry G.} \snm{Quinn}\thanksref{m2}\ead[label=e2]{barry.quinn@mq.edu.au}},
\author{\fnms{I. Vaughan L.} \snm{Clarkson}\thanksref{m3}\ead[label=e3]{v.clarkson@uq.edu.au}},
\author{\fnms{Bill} \snm{Moran}\thanksref{m4}\ead[label=e4]{wmoran@unimelb.edu.au}},
\and
\author{\fnms{Badri N.} \snm{Vellambi}\thanksref{m1}\ead[label=e5]{badri.vellambi@unisa.edu.au}}
\runauthor{R. McKilliam et al.}

@@

\begin{abstract}
Estimating the coefficients of a noisy polynomial phase signal is important in many fields including radar, biology and radio communications. One approach attempts to perform polynomial regression on the phase of the signal.  This is complicated by the fact that the phase is \emph{wrapped} modulo $2\pi$ and must be \emph{unwrapped} before regression can be performed. %A recent approach suggested by the authors is to perform the unwrapping in a least squares manner.  %It was shown how the resulting estimator could be posed as a \emph{nearest lattice point problem} in a specific lattice and could be solved using existing algorithms. 
%It was shown by Monte Carlo simulation that this produces a remarkably accurate estimator.  
In this paper we consider an estimator that performs phase unwrapping in a least squares manner.  We describe the asymptotic properties of this estimator, showing that it is strongly consistent and asymptotically normally distributed. %We hypothesise that the estimator produces very near maximum likelihood performance.
\end{abstract}

@@

Polynomial phase signals arise in numerous fields including radar, sonar, geophysics, speech analysis, biology, and radio communication~\cite{Hlawatsch_lin_quad_time_freq_spmag_1992,Ridleyspeechpolyphase1989, Suga_1975_bats_echolocation, Moss_2005echolocation}. In radar and sonar applications polynomial phase signals arise when acquiring radial velocity and acceleration (and higher order motion descriptors) of a target from a reflected signal, and also in continuous wave radar and low probability of intercept radar.  In biology, polynomial phase signals are used to describe the sounds emitted by bats and dolphins for echo location.

A polynomial phase signal of order $m$ is a function of the form
\[
s(t) = e^{2\pi j y(t)},
\]
where $j = \sqrt{-1}$, and $t$ is a real number, often representing time, and 
\[
y(t) = \tilde{\mu}_0 +\tilde{\mu}_1 t + \tilde{\mu}_2 t^2 + \dots \tilde{\mu}_m t^m
\]
is a polynomial of order $m$.  In practice the signal is typically sampled at discrete points in `time', $t$. In this paper we only consider uniform sampling, where the gap between consecutive samples is constant. In this case we can always consider the samples to be taken at some set of consecutive integers and our sampled polynomial phase signal is $s_n = s(n) = e^{2\pi j y(n)}$, where $n$ is an integer.  Of practical importance is the estimation of the coefficients $\tilde{\mu}_0, \dots, \tilde{\mu}_m$ from a number, say $N$, of consequetive observations of the noisy sampled signal
\begin{equation}\label{eq:Y_nsamplednoisey}
Y_n = \rho s_n + X_n,
\end{equation}
where $\rho$ is a real number greater than zero representing the (usually unknown) signal amplitude and $\{X_n, n \in \ints\}$ is a sequence complex noise variables. In order to ensure identifiability it is necessary to restrict the $m+1$ coefficients to a region of $m+1$ dimensional Euclidean space $\reals^{m+1}$ called an \emph{identifiable region}.  It was shown in \cite{McKilliam2009IndentifiabliltyAliasingPolyphase} that an identifiable region tessellates a particular $m+1$ dimensional lattice.  We discuss this in Section \ref{sec:ident_aliasing}.

@@

In this paper we consider the estimator that results from unwrapping the phase in a least squares manner.  We call this the \emph{least squares unwrapping estimator} (LSU).  It was shown in \cite{McKilliam2009asilomar_polyest_lattice} that the LSU estimator can be computed by finding a nearest point in a lattice~\cite{Agrell2002}, and Monte-Carlo simulations were used to show the LSU estimator's favourable statistical performance. %Moreover, it was noticed that the LSU appears to work correctly for coefficients anywhere in the identifiable region.  A drawback of the LSU estimator is that computing a nearest lattice point is, in general, computationally difficult.  In \cite{McKilliam2009asilomar_polyest_lattice}, two standard techniques were considered, the \emph{sphere decoder} \cite{Pohst_sphere_decoder_1981}, and \emph{Babai's nearest plane algorithm} \cite{Babai1986}. The sphere decoder was observed to have excellent statistical performance but can only be computed efficiently for $N<50$ \cite{Jalden2005_sphere_decoding_complexity}.  Computing the Babai point requires only $O(N^2)$ operations, but its statistical performance is comparatively poor at low signal to noise ratio (SNR). A major point of interest is that the lattices considered are not \emph{random}, and therefore may admit fast nearest point algorithms. 
%This has been studied in~\cite[Sec.~4.2]{McKilliam2010thesis} where polynomial time algorithms where found that compute the nearest point exactly.  Unfortantely, although polynomial time, the algorithm are still computationally very slow in practice. 
% Two main questions were raised in \cite{McKilliam2009asilomar_polyest_lattice}:
% \begin{enumerate}
% \vspace{-0.1cm}
% \item What are the asymptotic statistical properties of the LSU estimator?
% \vspace{-0.18cm}
% \item Can the structure of the lattice be exploited to obtain fast (exact or approximate) nearest point algorithms and thereby a computationally efficient estimator?
% \vspace{-0.1cm}
% \end{enumerate}
% The purpose of this paper is to answer the first question. We prove that the LSU estimator is strongly consistent and asymptotically normally distributed. 
%We also apply another known approximate nearest point algorithm, called the $K$-best method \cite{Zhan2006_K_best_sphere_decoder}, that we show provides near sphere decoder performance, but can be computed in a reasonable amount of time if $N$ is less than about 300.
In this paper we derive the asymptotic properties of the LSU estimator.  Under some assumptions about the distribution of the noise $X_1, \dots, X_N$, we show the estimator to be strongly consistent and asymptotically normally distributed.  Similar results were stated without proof in~\cite{McKilliam_polyphase_est_icassp_2011}.  Here, we give a proof.  The results here are also more general than in~\cite{McKilliam_polyphase_est_icassp_2011}, allowing for a wider class of noise distributions.

An interesting property is that the estimator of the $k$th polynomial phase coefficient converges almost surely to $\tilde{\mu}_k$ at rate $N^{-k}$.  This is perhaps not surprising, since it is the same rate observed in polynomial regression.  However, asserting that convergence at this rate occurs is not trivial.  For this purpose we make use of elementary results about the number of arithmetic progressions that fit inside subsets of $\{1,2,\dots,N\}$.  This proof technique appears to be novel, and may be useful in other problems.  The proof of asymptotic normality is complicated by the fact that the objective function corresponding with the LSU estimator is not differentiable everywhere.  Techniques developed by Huber~\cite{Huberrobustest1964,Huberbehavmaxnonstandard1967} and refined by Pollard~\cite{Pollard_new_ways_clts_1986,Pollard_asymp_empi_proc_1989} become useful here.  We believe the LSU estimator to be an interesting new application of these techniques.  We are hopeful that the proof techniques developed here will be useful for purposes other than polynomial phase estimation, and in particular other applications involving data that is `wrapped' in some sense.  Potential candidates are the phase wrapped images observed in synthetic aperture radar and magnetic resonance imaging~\cite{Nico_phaseunwrappingSAR_2000}.

@@

A \term{lattice},  $\Lambda$, is a discrete subset of points in $\reals^n$ such that
\[
   \Lambda = \{\xbf = \Bbf\ubf \mid \ubf \in \ints^d \}
\]
where $\Bbf \in \reals^{n \times d}$ is an $n \times d$ matrix of rank $d$, called the \term{generator matrix} or \term{basis matrix} or simply generator or basis.  If $n = d$ the lattice is said to be \emph{full rank}.  Lattices are discrete Abelian groups under vector addition.  They are subgroups of the Euclidean group $\reals^n$.  Lattices naturally give rise to \emph{tessellations} of $\reals^n$ by the specification of a set of coset representatives for the quotient $\reals^n / \Lambda$.  One choice for a set of coset representatives is a \emph{fundamental parallelepiped}, i.e., the parallelepiped generated by the columns of a generator matrix.  Another choice is based on the \emph{Voronoi cell}, i.e., those points from $\reals^n$ nearest (with respect to a given norm) to the lattice point at the origin.  In what follows when we refer to the Voronoi cell, we will also mean the Vornoi cell with respect to the Euclidean norm.  It is always possible to construct a \emph{rectangular} set of representatives, as the next proposition will show.  We will use these rectangular regions for describing the aliasing of polynomial phase signals in Section~\ref{sec:ident_aliasing}.  These rectangular regions will be important for the derivation of the asymptotic properties of the LSU estimator in Section~\ref{sec:least-squar-unwr}.

\begin{proposition}\label{prop:lattice-theory-constructing_a_rectangular_tesselating_region}
Let  $\Lambda$ be an $n$ dimensional lattice and $\Bbf \in \reals^{n\times n}$ be a generator matrix for $\Lambda$. Let $\Bbf = \Qbf\Rbf$ where $\Qbf$ is orthonormal and $\Rbf$ is upper triangular with elements $r_{ij}$.  Then the rectangular prism $\Qbf P$ where
\[
P = \prod_{k=1}^{n}{\left[-\frac{r_{kk}}{2}, \frac{r_{kk}}{2}\right)}
\]
is a set of coset representatives for $\reals^n / \Lambda$.
\end{proposition}
\begin{proof}
This result is well known~\cite[Chapter IX, Theorem IV]{Cassels_geom_numbers_1997}~\cite[Proposition 2.1]{McKilliam2010thesis}.  This result is for lattices with full rank.  A result in the general case can be obtained similarly, but is not required here.  
\end{proof}

@@

Given a polynomial $g(x) = a_0 + a_1x + \dots + a_m x^m$, let
\[
\coef(g) = \left[ \begin{array}{ccccc} a_0 & a_1 & a_2 & \dots & a_m, \end{array} \right]^\prime,
\]
where superscript $^\prime$ indicates the transpose, denote the column vector of length $m+1$ containing the coefficients of $g$.  If $y$ and $z$ differ by a polynomial from $\mathcal{Z}$ then we can write $y = z + p$ where $p \in \mathcal{Z}$ and then also $\coef(y) = \coef(z) + \coef(p)$.
%\footnote{In group theory terminology $\coef(\cdot)$ coupled with vector addition is called a \term{group homomorphism}.\index{group homomorphism}}. 
  Consider the set
\[
L_{m+1} = \{ \coef(p) \mid p \in \mathcal{Z} \},
\]
containing the coefficient vectors corresponding to the polynomials in $\mathcal{Z}$.  Since the integer valued polynomials are a basis for $\mathcal{Z}$,
\begin{align*}
L_{m+1} &= \{ \coef(c_0 p_0 + c_1p_1 + \dots + c_mp_m) \mid c_i \in \ints \} \\
&= \{ c_0 \coef(p_0) + \dots + c_m\coef(p_m) \mid c_i \in \ints \}.
\end{align*}
Let
\[
\Pbf = \left[ \begin{array}{cccc} \coef(p_0)& \coef(p_1)& \dots& \coef(p_m)  \end{array} \right]
\]
be the $m+1$ by $m+1$ matrix with columns given by the coefficients of the integer valued polynomials.  Then,
\[
L_{m+1} = \{ \xbf = \Pbf\ubf \mid \ubf \in \ints^{m+1} \}
\]
and it is clear that $L_{m+1}$ is an $m+1$ dimensional lattice.  That is, the set of coefficients of the polynomials from $\mathcal{Z}$ forms a lattice with generator matrix $\Pbf$. We can restate Theorem~\ref{thm:circpolysampledthm} as:
\begin{corollary}\label{cor:circpolysampledcoef}
Two polynomial phase signals $s(t) = e^{2\pi j y(t)}$  and $r(t) = e^{2\pi j z(t)}$ are aliases if and only if $\coef(y)$ and $\coef(z)$ differ by a lattice point in $L_{m+1}$.
\end{corollary}

@@

We now describe the least squares unwrapping (LSU) estimator of the polynomial coefficients. Recall that we desire to estimate the coefficients $\tilde{\mu}_0, \dots, \tilde{\mu}_m$ from the noisy samples $Y_1, \dots, Y_N$ given in~\eqref{eq:Y_nsamplednoisey}.  We take the complex argument of the $Y_n$ and divide by $2\pi$ to obtain
\begin{equation}\label{eq:noise_circ_poly}
\Theta_n = \frac{\angle{Y_n}}{2\pi} = \fracpart{ \Phi_n + y(n) }
\end{equation}
where $\angle{z}$ denotes the complex argument of the complex number $z$, and 
\[
\Phi_n = \frac{1}{2\pi}\angle(1 + \rho^{-1}s_n^{-1}X_n)
\] 
are random variables representing the phase noise induced by the $X_n$~\cite{Tretter1985,Quinn2009_dasp_phase_only_information_loss}.  If the distribution of $X_n$ is circularly symmetric (i.e., the angle $\angle X_n$ is uniformly distributed on $[-\pi, \pi)$ and is independent of the magnitude $\abs{X_n}$) then the distribution of $\Phi_n$ is the same as the distribution of $\tfrac{1}{2\pi}\angle(1 + \rho^{-1}X_n)$.  If the $X_1, \dots, X_N$ are circularly symmetric and identically distributed, then $\Phi_1, \dots, \Phi_n$ are also identically distributed.

Let $\mubf$ be the vector $[\mu_0, \mu_1, \dots, \mu_m]$ and put,
\begin{equation} \label{eq:sumofsquaresfunction}
SS(\mubf) = \sum_{n=1}^{N}\fracpart{  \Theta_{n} - \sum_{k = 0}^{m}{\mu_k n^k} }^{2}.
\end{equation}
The least squares unwrapping estimator is defined as those coefficients $\widehat{\mu}_0, \dots, \widehat{\mu}_m$ that minimise $SS$ over the identifiable region $B$. That is, the LSU estimator is,
\begin{equation}\label{eq:hatmubfLSUdefn}
\widehat{\mubf} = \arg \min_{\mubf \in B} SS(\mubf). 
\end{equation}

It is shown in~\cite{McKilliam2009asilomar_polyest_lattice} and \cite[Sec~8.1]{McKilliam2010thesis} how this minimisation problem can be posed as a nearest lattice point problem, and we briefly discuss this here.  Write $SS$ as
\[
SS(\mubf) = \sum_{n=1}^{N}{ \left( \Theta_{n} - W_n - \sum_{k = 0}^{m}{\mu_k n^k} \right)^2 },
\]
where $W_n = \round{\Theta_{n} - \sum_{k = 0}^{m}{\mu_k n^k} }$ for $n = 1, \dots, N$ are integers called \emph{wrapping variables}.  If we consider $W_1,\dots,W_N$ as nuisance parameters to be estimated then $SS$ can be written as a function of both $\mu_0,\dots,\mu_m$ and $W_1,\dots,W_N$.  The least squares unwrapping estimator is then found by jointly minimising over $\mu_0,\dots,\mu_m$ and $W_1,\dots,W_N$.   This joint minimisation problem can be solved by computing a nearest point in a lattice.  To see this write $SS$ as a function of both $\mu_0,\dots,\mu_m$ and $W_1,\dots,W_N$ using vectors,
\begin{equation}\label{eq:SS_in_vector_form_with_u_nuisance}
SS(\mubf, \wbf) = \| \thetabf - \Xbf\mubf - \wbf \|^2,
\end{equation}
where $\|\cdot\|^2$ denotes the squared Euclidean norm, and $\Xbf$ is the rectangular Vandermonde matrix
\begin{equation}\label{eq:Xvandermonde}
\Xbf = \left[ \begin{array}{ccccc} 
1 & 1 & 1 & \cdots & 1 \\
1 & 2 & 4 & \cdots & 2^m \\
\vdots & \vdots  & \vdots & \ddots & \vdots \\
1 & N & N^2 & \cdots & N^m \\
\end{array} \right],
\end{equation}
and where the column vectors $\wbf = [W_1, \dots, W_N]^\prime$ and $\thetabf = [\Theta_1, \dots, \Theta_N]^\prime$.   For fixed $\wbf$ the minimiser of $SS(\mubf, \wbf)$ with respect to $\mubf$ is
\begin{equation} \label{eq:hat_theta_cal}
\mubf(\wbf) = (\Xbf^\prime\Xbf)^{-1}\Xbf^\prime( \thetabf - \wbf) = \Xbf^+( \thetabf - \wbf),
\end{equation}
where $\Xbf^+ = (\Xbf^\prime\Xbf)^{-1}\Xbf^\prime$ is the Moore-Penrose pseudoinverse of $\Xbf$.

We may thus compute $\widehat{\mubf}$ as $\widehat{\mubf} = \mu(\widehat{\wbf})$, where $\widehat{\wbf}$ is the minimiser of
\[
SS(\wbf) = SS(\mubf(\wbf),\wbf) = \| \Qbf\thetabf - \Qbf\wbf \|^2,
\]
where $\Qbf = \Ibf - \Xbf\Xbf^+$ and $\Ibf$ is the $N \times N$ identity matrix\footnote{We have slightly abused notation here by reusing $SS$. This should not cause any confusion as $SS(\mubf)$ and $SS(\mubf,\wbf)$ and $SS(\wbf)$ are easily told apart by their inputs.}.

Let $\Lambda$ be the lattice generated by $\Qbf$, that is $\Lambda = \{\Qbf\xbf \mid \xbf \in \ints^N \}$\footnote{The $N \times N$ projection matrix $\Qbf$ does not have full rank, it has rank $N-m-1$, and for this reason it is not strictly a generator matrix.  However, a generator can be constructed by removing $m+1$ consecutive columns from $\Qbf$. In our implementation the last $m+1$ columns are removed.}.  It follows that $\Qbf \wbf$ is a lattice point in $\Lambda$ and that minimising $SS(\wbf)$ is equivalent to finding the nearest lattice point in $\Lambda$ to $\Qbf\thetabf$.  Denote this point by $\Qbf\widehat{\wbf}$.  Now the estimate $\widehat{\mubf}$ is given by substituting $\widehat{\wbf}$ for $\wbf$ in~\eqref{eq:hat_theta_cal}.  After this procedure it is possible that the $\widehat{\mubf}$ obtained is not in the identifiable region but is instead an aliased version of the desired estimate. This can be resolved by computing $\dealias(\widehat{\mubf})$.  There exist algorithms that can compute a nearest point in $\Lambda$ that require a number of operations that is polynomial in $N$~\cite[Sec 4.3]{McKilliam2010thesis}.  Although polynomial in complexity, these algorithms are not fast in practice.  The existence of practically fast nearest point algorithms for these lattices is an interesting open problem.

@@

 \begin{lemma} \label{lem:S(h,K)size}
 Let $K \subseteq W$.  Then 
\[
|A(h,K)| \geq N - mh - (N - |K|)(m+1).
\]
 \end{lemma}
 \begin{proof}
 Note that $|A(h,W)| = N - mh$.  The proof follows by starting with $A(h,W)$ and applying Lemma~\ref{lem:S(h,G/r)size} 
\[
|W|-|K| = N - |K|
\] 
times. That is, $K$ can be constructed by removing $N - |K|$ elements from $W$ and this removes at most $(N - |K|)(m+1)$ arithmetic progressions from $A(h,W)$.
 \end{proof}

@@

\section{An empirical process result}

>>>>

\begin{aug}
\author{\fnms{Robby G.} \snm{McKilliam}\thanksref{m1}\ead[label=e1]{robby.mckilliam@unisa.edu.au}},
\author{\fnms{Barry G.} \snm{Quinn}\thanksref{m2}\ead[label=e2]{barry.quinn@mq.edu.au}},
\author{\fnms{I. Vaughan L.} \snm{Clarkson}\thanksref{m3}\ead[label=e3]{v.clarkson@uq.edu.au}},
\author{\fnms{Bill} \snm{Moran}\thanksref{m4}\ead[label=e4]{wmoran@unimelb.edu.au}},
\and
\author{\fnms{Badri N.} \snm{Vellambi}\thanksref{m1}\ead[label=e5]{badri.vellambi@unisa.edu.au}}
\runauthor{R. McKilliam et. al.}

@@

\begin{abstract}
Estimating the coefficients of a noisy polynomial phase signal is important in fields including radar, biology and radio communications. One approach attempts to perform polynomial regression on the phase of the signal.  This is complicated by the fact that the phase is \emph{wrapped} modulo $2\pi$ and must be \emph{unwrapped} before regression can be performed. %A recent approach suggested by the authors is to perform the unwrapping in a least squares manner.  %It was shown how the resulting estimator could be posed as a \emph{nearest lattice point problem} in a specific lattice and could be solved using existing algorithms. 
%It was shown by Monte Carlo simulation that this produces a remarkably accurate estimator.  
In this paper we consider an estimator that performs phase unwrapping in a least squares manner.  We describe the asymptotic properties of this estimator, showing that it is strongly consistent and asymptotically normally distributed. %We hypothesise that the estimator produces very near maximum likelihood performance.
\end{abstract}

@@

Polynomial phase signals arise in fields including radar, sonar, geophysics, speech analysis, biology, and radio communication~\cite{Hlawatsch_lin_quad_time_freq_spmag_1992,Ridleyspeechpolyphase1989, Suga_1975_bats_echolocation, Moss_2005echolocation}. In radar and sonar applications polynomial phase signals arise when acquiring radial velocity and acceleration (and higher order motion descriptors) of a target from a reflected signal, and also in continuous wave radar and low probability of intercept radar.  In biology, polynomial phase signals are used to describe the sounds emitted by bats and dolphins for echo location.

A polynomial phase signal of order $m$ is a function of the form
\[
s(t) = e^{2\pi j y(t)},
\]
where $j = \sqrt{-1}$, and $t$ is a real number, often representing time, and 
\[
y(t) = \tilde{\mu}_0 +\tilde{\mu}_1 t + \tilde{\mu}_2 t^2 + \dots \tilde{\mu}_m t^m
\]
is a polynomial of order $m$.  In practice the signal is typically sampled at discrete points in `time', $t$. In this paper we only consider uniform sampling, where the gap between consecutive samples is constant. In this case we can always consider the samples to be taken at some set of consecutive integers and our sampled polynomial phase signal is $s_n = s(n) = e^{2\pi j y(n)}$, where $n$ is an integer.  Of practical importance is the estimation of the coefficients $\tilde{\mu}_0, \dots, \tilde{\mu}_m$ from a number, say $N$, of consecutive observations of the noisy sampled signal
\begin{equation}\label{eq:Y_nsamplednoisey}
Y_n = \rho s_n + X_n,
\end{equation}
where $\rho$ is a real number greater than zero representing the (usually unknown) signal amplitude and $\{X_n, n \in \ints\}$ is a sequence of complex noise variables. In order to ensure identifiability it is necessary to restrict the $m+1$ coefficients to a region of $m+1$ dimensional Euclidean space $\reals^{m+1}$ called an \emph{identifiable region}.  It was shown by some of the authors \cite{McKilliam2009IndentifiabliltyAliasingPolyphase} that an identifiable region tessellates a particular $m+1$ dimensional lattice.  We discuss this in Section \ref{sec:ident_aliasing}.

@@

In this paper we consider the estimator that results from unwrapping the phase in a least squares manner.  We call this the \emph{least squares unwrapping} (LSU) estimator.  It was shown by some of the authors \cite[Sec. 8.1]{McKilliam2010thesis}\cite{McKilliam2009asilomar_polyest_lattice} that the LSU estimator can be computed by finding a nearest point in a lattice~\cite{Agrell2002}, and Monte-Carlo simulations were used to show the LSU estimator's favourable statistical performance. %Moreover, it was noticed that the LSU appears to work correctly for coefficients anywhere in the identifiable region.  A drawback of the LSU estimator is that computing a nearest lattice point is, in general, computationally difficult.  In \cite{McKilliam2009asilomar_polyest_lattice}, two standard techniques were considered, the \emph{sphere decoder} \cite{Pohst_sphere_decoder_1981}, and \emph{Babai's nearest plane algorithm} \cite{Babai1986}. The sphere decoder was observed to have excellent statistical performance but can only be computed efficiently for $N<50$ \cite{Jalden2005_sphere_decoding_complexity}.  Computing the Babai point requires only $O(N^2)$ operations, but its statistical performance is comparatively poor at low signal to noise ratio (SNR). A major point of interest is that the lattices considered are not \emph{random}, and therefore may admit fast nearest point algorithms. 
%This has been studied in~\cite[Sec.~4.2]{McKilliam2010thesis} where polynomial time algorithms where found that compute the nearest point exactly.  Unfortantely, although polynomial time, the algorithm are still computationally very slow in practice. 
% Two main questions were raised in \cite{McKilliam2009asilomar_polyest_lattice}:
% \begin{enumerate}
% \vspace{-0.1cm}
% \item What are the asymptotic statistical properties of the LSU estimator?
% \vspace{-0.18cm}
% \item Can the structure of the lattice be exploited to obtain fast (exact or approximate) nearest point algorithms and thereby a computationally efficient estimator?
% \vspace{-0.1cm}
% \end{enumerate}
% The purpose of this paper is to answer the first question. We prove that the LSU estimator is strongly consistent and asymptotically normally distributed. 
%We also apply another known approximate nearest point algorithm, called the $K$-best method \cite{Zhan2006_K_best_sphere_decoder}, that we show provides near sphere decoder performance, but can be computed in a reasonable amount of time if $N$ is less than about 300.
In this paper we derive the asymptotic properties of the LSU estimator.  Under some assumptions about the distribution of the noise $X_1, \dots, X_N$, we show the estimator to be strongly consistent and asymptotically normally distributed.  Similar results were stated without proof in~\cite{McKilliam_polyphase_est_icassp_2011}.  Here, we give a proof.  The results here are also more general than in~\cite{McKilliam_polyphase_est_icassp_2011}, allowing for a wider class of noise distributions.

An interesting property is that the estimator of the $k$th polynomial phase coefficient converges almost surely to $\tilde{\mu}_k$ at rate $N^{-k}$.  This is perhaps not surprising, since it is the same rate observed in polynomial regression.  However, asserting that convergence at this rate occurs in the polynomial phase setting is not trivial.  For this purpose we make use of an elementary result about the number of arithmetic progressions contained inside subsets of $\{1,2,\dots,N\}$.  %This proof technique appears to be novel, and may be useful in other problems.  
The proof of asymptotic normality is complicated by the fact that the objective function corresponding with the LSU estimator is not differentiable everywhere.  Techniques developed by Huber~\cite{Huberrobustest1964,Huberbehavmaxnonstandard1967} and refined by Pollard~\cite{Pollard_new_ways_clts_1986,Pollard_asymp_empi_proc_1989} and results from the literature on hyperplane arrangements~\cite{Chazelle_discrepency_method_2000,Matousek_lect_disc_geom_2002} become useful here.  We are hopeful that the proof techniques developed here will be useful for purposes other than polynomial phase estimation, and in particular other applications involving data that is `wrapped' in some sense.  Potential candidates are the phase wrapped images observed in synthetic aperture radar and magnetic resonance imaging~\cite{Nico_phaseunwrappingSAR_2000}.

@@

A \term{lattice},  $\Lambda$, is a discrete subset of points in $\reals^n$ such that
\[
   \Lambda = \{\xbf = \Bbf\ubf \mid \ubf \in \ints^d \}
\]
where $\Bbf \in \reals^{n \times d}$ is an $n \times d$ matrix of rank $d$, called the \term{generator matrix} or \term{basis matrix} or simply generator or basis.  If $n = d$ the lattice is said to be \emph{full rank}.  Lattices are discrete Abelian groups under vector addition.  They are subgroups of the Euclidean group $\reals^n$.  Lattices naturally give rise to \emph{tessellations} of $\reals^n$ by the specification of a set of coset representatives for the quotient $\reals^n / \Lambda$.  One choice for a set of coset representatives is a \emph{fundamental parallelepiped}, i.e., the parallelepiped generated by the columns of a generator matrix.  Another choice is based on the \emph{Voronoi cell}, i.e., those points from $\reals^n$ nearest (with respect to the Euclidean norm here ) to the lattice point at the origin.  It is always possible to construct a \emph{rectangular} set of representatives, as the next proposition will show.  We will use these rectangular regions for describing the aliasing properties of polynomial phase signals in Section~\ref{sec:ident_aliasing}.  These rectangular regions will be important for the derivation of the asymptotic properties of the LSU estimator in Section~\ref{sec:least-squar-unwr}.

\begin{proposition}\label{prop:lattice-theory-constructing_a_rectangular_tesselating_region}
Let  $\Lambda$ be an $n$ dimensional lattice and $\Bbf \in \reals^{n\times n}$ be a generator matrix for $\Lambda$. Let $\Bbf = \Qbf\Rbf$ where $\Qbf$ is orthonormal and $\Rbf$ is upper triangular with elements $r_{ij}$.  Then the rectangular prism $\Qbf P$ where $P = \prod_{k=1}^{n}{[-\tfrac{r_{kk}}{2}, \frac{r_{kk}}{2})}$ is a set of coset representatives for $\reals^n / \Lambda$.
\end{proposition}
\begin{proof}
This result is well known~\cite[Chapter IX, Theorem IV]{Cassels_geom_numbers_1997}~\cite[Proposition 2.1]{McKilliam2010thesis}.  This result is for lattices with full rank.  A result in the general case can be obtained similarly, but is not required here.  
\end{proof}

@@

Given a polynomial $g(x) = a_0 + a_1x + \dots + a_m x^m$, let
\[
\coef(g) = \left[ \begin{array}{ccccc} a_0 & a_1 & a_2 & \dots & a_m, \end{array} \right]^\prime
\]
denote the column vector of length $m+1$ containing the coefficients of $g$.  We use superscript $^\prime$ to indicate the vector or matrix transpose.  If $y$ and $z$ differ by a polynomial from $\mathcal{Z}$ then $y = z + p$ where $p \in \mathcal{Z}$ and also $\coef(y) = \coef(z) + \coef(p)$.
%\footnote{In group theory terminology $\coef(\cdot)$ coupled with vector addition is called a \term{group homomorphism}.\index{group homomorphism}}. 
  Consider the set
\[
L_{m+1} = \{ \coef(p) \mid p \in \mathcal{Z} \}
\]
containing the coefficient vectors corresponding to the polynomials in $\mathcal{Z}$.  Since the integer valued polynomials are a basis for $\mathcal{Z}$,
\begin{align*}
L_{m+1} &= \{ \coef(c_0 p_0 + c_1p_1 + \dots + c_mp_m) \mid c_i \in \ints \} \\
&= \{ c_0 \coef(p_0) + \dots + c_m\coef(p_m) \mid c_i \in \ints \}.
\end{align*}
Let
\[
\Pbf = \left[ \begin{array}{cccc} \coef(p_0)& \coef(p_1)& \dots& \coef(p_m)  \end{array} \right]
\]
be the $m+1$ by $m+1$ matrix with columns given by the coefficients of the integer valued polynomials.  Then,
\[
L_{m+1} = \{ \xbf = \Pbf\ubf \mid \ubf \in \ints^{m+1} \}
\]
and it is clear that $L_{m+1}$ is an $m+1$ dimensional lattice.  That is, the set of coefficients of the polynomials from $\mathcal{Z}$ forms a lattice with generator matrix $\Pbf$. We can restate Theorem~\ref{thm:circpolysampledthm} as:
\begin{corollary}\label{cor:circpolysampledcoef}
Two polynomial phase signals $s(t) = e^{2\pi j y(t)}$  and $r(t) = e^{2\pi j z(t)}$ are aliases if and only if $\coef(y)$ and $\coef(z)$ differ by a lattice point in $L_{m+1}$.
\end{corollary}

@@

We now describe the least squares unwrapping (LSU) estimator of the polynomial coefficients. Recall that we desire to estimate the coefficients $\tilde{\mu}_0, \dots, \tilde{\mu}_m$ from the noisy samples $Y_1, \dots, Y_N$ given in~\eqref{eq:Y_nsamplednoisey}.  We take the complex argument of the $Y_n$ and divide by $2\pi$ to obtain
\begin{equation}\label{eq:noise_circ_poly}
\Theta_n = \frac{\angle{Y_n}}{2\pi} = \fracpart{ \Phi_n + y(n) }
\end{equation}
where $\angle$ denotes the complex argument (or phase), and 
\[
\Phi_n = \frac{1}{2\pi}\angle(1 + \rho^{-1}s_n^{-1}X_n)
\] 
are random variables representing the `phase noise' induced by the $X_n$.  If the distribution of $X_n$ is circularly symmetric (the phase $\angle X_n$ is uniformly distributed on $[-\pi, \pi)$ and is independent of the magnitude $\abs{X_n}$) then the distribution of $\Phi_n$ is the same as the distribution of $\tfrac{1}{2\pi}\angle(1 + \rho^{-1}X_n)$.  If $X_1, \dots, X_N$ are circularly symmetric and identically distributed, then $\Phi_1, \dots, \Phi_n$ are also identically distributed.

Let $\mubf$ be the vector $[\mu_0, \mu_1, \dots, \mu_m]$ and put,
\begin{equation} \label{eq:sumofsquaresfunction}
SS(\mubf) = \sum_{n=1}^{N}\fracpart{  \Theta_{n} - \sum_{k = 0}^{m}{\mu_k n^k} }^{2}.
\end{equation}
The least squares unwrapping estimator is defined as those coefficients $\widehat{\mubf} = [\widehat{\mu}_0, \dots, \widehat{\mu}_m]$ that minimise $SS$ over the identifiable region $B$, i.e., the LSU estimator is,
\begin{equation}\label{eq:hatmubfLSUdefn}
\widehat{\mubf} = \arg \min_{\mubf \in B} SS(\mubf). 
\end{equation}

It is shown in~\cite{McKilliam2009asilomar_polyest_lattice} how this minimisation problem can be posed as that of computing a nearest lattice point in a particular lattice. Polynomial time algorithms that compute the nearest point are described in~\cite[Sec 4.3]{McKilliam2010thesis}.  Although polynomial in complexity, these algorithms are not fast in practice.  The existence of practically fast nearest point algorithms for these lattices is an interesting open problem.  We focus on the asymptotic statistical properties of the LSU estimator, rather than computational aspects in this paper.

@@

 \begin{lemma} \label{lem:S(h,K)size}
 Let $K \subseteq W$.  Then $|A(h,K)| \geq N - mh - (N - |K|)(m+1)$.
 \end{lemma}
 \begin{proof}
 Note that $|A(h,W)| = N - mh$.  The proof follows by starting with $A(h,W)$ and applying Lemma~\ref{lem:S(h,G/r)size} 
\[
|W|-|K| = N - |K|
\] 
times. That is, $K$ can be constructed by removing $N - |K|$ elements from $W$ and this removes at most $(N - |K|)(m+1)$ arithmetic progressions from $A(h,W)$.
 \end{proof}

@@

\section{A uniform law of large numbers} \label{app:uniform-law-large}

During the proof of strong consistency we made use of the fact that
\begin{equation}~\label{eq:SNVNunifmlln2}
\sup_{\lambdabf \in B}\sabs{S_N(\lambdabf) - V_N(\lambdabf)} \rightarrow 0
\end{equation}
almost surely as $N\rightarrow\infty$, where $V_N(\lambdabf) = \expect S_N(\lambdabf)$.  We prove this result here. Put $D_N(\lambdabf) = S_N(\lambdabf) - V_N(\lambdabf)$.  By Lemma~\ref{lem:vn},
\[
\sum_{N=1}^\infty \prob \left\{ \sup_{\lambdabf \in B}\abs{ D_N(\lambdabf) } > \epsilon \right\} < \infty
\]
for any $\epsilon > 0$.  Thus~\eqref{eq:SNVNunifmlln2} follows from the Borel-Cantelli lemma.  In what follows we use order notation in the standard way, that is, for functions $h$ and $g$, we write $h(N) = O(g(N))$ to mean that there exists a constant $K > 0$ and a finite $N_0$ such that $h(N) \leq K g(N)$ for all $N > N_0$.
